Intestinal microbiome dysbiosis increases Mycobacteria pulmonary colonization in mice by regulating the Nos2-associated pathways

Version 1

Abstract

Abstract

Increasing researches reveal gut microbiota was associated with the development of tuberculosis. How to prevent or reduce Mycobacterium tuberculosis colonization in the lungs is a key measure to prevent tuberculosis. However, the data on gut microbiota preventing Mycobacterium colonization in the lungs were scarce. Here, we established the clindamycin-inducing intestinal microbiome dysbiosis and fecal microbial transplantation models in mice to identify the effect of gut microbiota on the colonization of Mycobacterium in the mouse lungs and to explore its potential mechanisms. The results showed that clindamycin treatment altered the diversity and composition of the intestinal bacterial and fungal microbiome, weakened the trans-kingdom network interactions between bacteria and fungi, and induced gut microbiome dysbiosis in the mice. Gut microbiota dysbiosis increases intestinal permeability and enhances the susceptibility of Mycobacterium colonization in the lungs of mice. The potential mechanisms were gut microbiota dysbiosis altered the lung transcriptome and increased Nos2 expression through the “gut-lung axis”. Nos2 high expression disrupts the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1 in the cells, and resulting in promoting Mycobacteria colonization in the lungs of mice. The present study raises a potential strategy for reducing the risks of Mycobacteria infections and transmission by regulating the gut microbiome balance.

Introduction

Tuberculosis (TB) is a chronic infectious disease caused by Mycobacterium tuberculosis (Mtb), with approximately 10 million people being infected each year and 1.6 million deaths in 20211,2. The epidemic of TB has brought heavy economic and social burdens to countries around the world. According to the World Health Organization (WHO) global TB report 2022, the global spending on essential tuberculosis services for TB diagnosis, treatment, and prevention has exceeded $6.4 billion3. Besides, Drug-resistant forms of Mtb are currently on course to be the world’s deadliest pathogens, responsible for a quarter of deaths due to antimicrobial resistance1. The prevalence of drug-resistant Mtb is also a major threat to the control of TB. Hence, it has been an urgent task for researchers to develop more effective measures for the prevention and treatment of TB.

Recent advances in gut microbiota explorations have led to improved knowledge of the “gut-lung axis”. Some studies have shown that intestinal microbiota dysbiosis may affect the occurrence and development of respiratory diseases including asthma, chronic obstructive pulmonary disease, and respiratory pathogens infection et al by the gut-lung axis4–6. The documents about the relationship between gut microbiota and TB have increased in recent years, and increasing evidence supports the presence of intestinal microbiota dysbiosis in TB patients. A case-control study of pulmonary TB in pediatric patients showed that the abundance of Prevotellaceae and Enterococcaceae were increased, and the abundance of Oscillospiraceae and Bifidobacteriaceae were decreased in TB patients compared with healthy control subjects7. Maji and colleagues have found that the levels of Provetalla decreased and Bacteroides increased in TB patients8. Recently, our teams’ study in active TB patients showed the abundance of Bacteroidetes significantly altered compared with health subjects. Another study also showed the phylum of Bacteroidetes altered in the feces samples of patients with recurrent TB9. All these researches indicated that Mtb infection can lead to the alteration of intestinal microbiota composition.

However, the causal relation between Mtb infection and intestinal microbiota dysbiosis has not been fully elucidated. Some factors including alcohol, smoking, HIV infection, and diabetes have been proven to cause gut microbiota dysbiosis, and these also are important risk factors for TB10. A recent study has shown that the mice fed with a high-fat diet-induced gut microbiota dysbiosis increased the risk of developing active TB in mice11. Another study found Firmicutes/Bacteroidetes ratio increased in a murine model of type 2 diabetes induced by an energy-dense diet, and increased the susceptibility to Mtb infection12. These findings support a viewpoint that the changes in the gut microbiome are a contributing factor in Mtb infection pathogenesis and its clinical presentation. Many documents supposed that gut microbiota dysbiosis could modulate immunity and inflammation reaction at distal sites such as the lungs, reduce colonization resistance by external pathogens, and promote TB susceptibility10,13. Hence, modulation of the gut microbiota and balance of the gut-lung axis was a potential avenue for TB treatment and management.

Based on our previous studies and other reports, we found the effect of Mtb infection on the gut Bacteroidetes was most significant. Hence, in the present study, clindamycin, an antibiotic that selectively disorders anaerobic Bacteroidetes14, was used to establish a mouse model of intestinal microbiota dysbiosis. We uncovered the important role of gut microbiome dysbiosis affecting Mycobacteria colonization in the mouse lungs using the technology of fecal microbiota transplantation (FMT). Furthermore, based on the “gut-lung axis” theory, we performed a transcriptomic analysis of the mice lungs after Mycobacteria infection. We aim to explore the potential role and mechanism of gut microbiome dysbiosis enhancing Mycobacteria colonization in the lungs of mice.

Results

Clindamycin increases Mycobacterium smegmatis colonization in the lungs of mice

According to the Figure 1a procedure, we established a clindamycin-treated mice model to assess the effects of clindamycin-associated gut dysbiosis on MS colonization in the lungs of mice. The results showed that the mice with clindamycin treated presented a significantly higher colonization of MS in the lungs than that of control mice (Figure 1b). The size of the cecum of CL-treated mice was markedly more dilated compared with the control groups (Figure 1c). To assess the effect of CL on gut mucosal damage and permeability, the intestinal fatty acid binding protein (i-FABP, a marker of enterocyte death) and lipopolysaccharides (LPS, an endotoxin, as a marker of gut permeability) in serum were investigated with enzyme-linked immunosorbent assay (ELISA) technology. The results showed that the levels of i-FABP and LPS were significantly higher in the CL group than in the control group (CT) (Figure 1d and 1e). Pathological sections of intestinal tissue showed the intestinal epithelial tissues of mice with no obvious alteration after clindamycin treatment vs the CON group (Figure S1). All these results indicated that CL treatment damages the enterocyte, increases gut permeability, promotes the fermentation of cecum contents, and increases MS colonization in the lungs of mice.

Clindamycin increases MS pulmonary colonization in mice.

(a) The experimental procedures of clindamycin-inducing gut microbiota dysbiosis model in mice. (b) The colonization of MS in the lungs of mice after infection at 72 h. (c) The length of cecum after clindamycin treatment. (d)The level of iFABP in serum. (e) LPS concentration in serum. CL: clindamycin-treatment group, CON: control group. MS: Mycobacterium smegmatis, iFABP: intestinal fatty acid binding protein, LPS: lipopolysaccharides. *P<0.05, **P<0.01, *** P<0.001.

Altered the diversity and composition of the gut microbiome in mice

To assess the effect of CL treatment on the gut microbiome of mice, we investigated the alteration of gut microbiota (bacterial) and mycobiota (fungi) by 16S rRNA and ITS2 amplicon sequencing, respectively. For the gut bacterial analysis, the α-diversity was significantly decreased in the CL group compared with the CON group (Figure 2a). The β-diversity based on the weighted UniFrac distance showed that the CL group samples were separated from the CON group (R=0.8488, P=0.001) (Figure 2b). It indicated that the intestinal bacterial microbial community structure was a significant difference between the CL-treated mice and the control mice. At the phylum level, we observed a remarkable decrease of Bacteroidota and a significant increase of Firmicutes and Proteobacteria in the CL group compared to the control group (Figure 2c, Table S1). At the genus level, the CL group had a significantly higher relative abundance of Bacteroides, Lactobacillus, Escherichia-Shigella, and Faecalibaculum, and a lower abundance of Dubosiella, Alloprevotella, Akkermansia, and Bifidobacterium compared with that in the CON group (Figure 2d, Table S2).

Clindamycin altered the diversity and composition of the gut microbiome in mice.

(a) Chao1 and Inverse Simpson index of gut microbiota. (b) PCoA analysis of gut microbiota based on weighted Unifrac distance. (c) The composition distribution of gut microbiome (bacteria and fungi) at the phylum level. (d) The LEfSe analysis of the differentially abundant gut microbiota between the CL group and CON group at the genera level (LDA>4, P<0.05). CL: clindamycin-treatment group, CON: control group. **P<0.01, **** P<0.0001.

For the fungal mycobiota analysis, the Chao 1 and inverse Simpson index of the CL group was significantly increased compared to that of the CON group (Figure 2a). The PCoA showed that the samples were significantly distinguished between two groups (R=0.6038, P=0.001) (Figure 2b), which suggested that the community of intestinal mycobiota in the two groups is different. At the phylum level, the fungal microbiota was dominated by the Ascomycota and Basidiomycota in two groups, and Ascomycota showed no obvious differences between the two groups. However, the relative abundance of Basidiomycota significantly increases in the CL group compared with the CON group (Figure 2c. Table S3). The LEfSe analysis at the genus level showed that the relative abundance of Aspergillus, Penicillium, Cladosporium, and Alternaria was enriched in the CL group vs the CON group (Figure 2d, Table S4). These results are consistent with the Wilcoxon rank-sum test analysis (Figure S2). Taken together, all these data indicated that CL treatment altered the diversity and composition of the intestinal microbiome (including bacterial and fungi) in the mice and induced gut microbiome dysbiosis.

Clindamycin weakens the trans-kingdom networks construction of gut bacterial and fungi

To evaluate the effect of clindamycin treatment on gut microbiome balance, we analyzed the fungi-to-bacteria species ratio with ITS/16S. The results show that the ITS/16S ratio of the CL group was significantly increased compared with the CON group (Figure 3a). We also assessed the alterations in the ratio of the dominant phyla in the gut microbiome. The results showed that the ratio of Firmicutes/Bacteroidota in the CL group was increased, and Ascomycota/Basidiomycota was significantly decreased compared with that of the CON group (Figure 3b and 3c). These results revealed that clindamycin treatment not only disturbed the intestinal equilibrium between bacteria and fungi but also destroyed the balance among the dominant bacteria (or fungi). Then, to assess the interplay between bacteria and fungi, we performed the trans-kingdom networks analysis at the genus level. The results showed that the trans-kingdom networks were altered in the CL group vs that in the CON group. In the control group, the bacteria and fungi were closely related to each other, gathering in a cluster and forming a more complex network (Figure 3d). There was a total of 154 nodes including 45 fungi and 109 bacteria, and 810 edges, and the relative connectedness was 5.26 (Table 1). However, the complexity of the trans-kingdom network in the CL group was dramatically decreased, and the interplay between bacteria and fungi was also weakened (Figure 3e). A total of 110 nodes including 62 fungi and 48 bacteria gathered into clusters in this network. The edges decreased to 357 and the relative connectedness was 3.25 in the CL group network (Table 1). We also found that the interplay ratio between fungi in the CL group was 52.10%, which was much higher than that of the CON group (7.53%) (Table 1). All these results indicated that clindamycin treatment not only altered the proportion of bacteria and fungi, but also weakened the trans-kingdom networks, changed the interactions between bacteria and fungi, and resulted in gut microbiome dysbiosis.

Clindamycin weakens the trans-kingdom network construction of gut bacteria and fungi.

(a) The ratio of ITS2/16S at the genus level. (b) The relative abundant ratio of Firmicutes/Bacteroidota. (c) The relative abundant ratio of Ascomycota/Basidiomycota. (d) The trans-kingdom correlation networks of the CON at the genus level. (e) The trans-kingdom correlation networks of the CL at the genus level. CL: clindamycin-treatment group, CON: control group. *P<0.05; ***P<0.001.

Fecal microbial transfer from Clindamycin-treated mice donors to antibiotics-treated conventional mice enhances MS colonization

To further verify the relationship of gut microbiota dysbiosis with the susceptibility of MS infection in the mice’s lungs, we performed an FMT model. The experimental design for the FMT is shown in the Figure 4a. First, we analyzed the α-diversity of gut bacterial and fungal microbiota after FMT with the Chao 1 index and inverse Simpson index. The results show that the α-diversity of gut bacterial microbiota was significantly decreased, and the α-diversity of gut fungal microbiota was significantly increased in the CL-FMT group compared with that in the CON-FMT group (Figure S3). Then we performed PCoA analysis to distinguish the gut microbiome alteration between CL- and CON-recipient mice. The results show a significant separation of gut bacterial and fungal microbiota between CT- and AD-recipient samples (p=0.007 and p=0.042, Figure 4b). The fungal-to-bacterial species ratio was significantly increased in the CL-recipient group vs the CON-recipient group (p<0.05, Figure 4c). To calculate the changes of the dominant phyla in the gut microbiome, the results show that Firmicutes/Bacteroidota in the CL-FMT group was decreased, and Ascomycota/Basidiomycota was no obvious different compared with that of the CON-FMT group (Figure 4d and 4e). The trans-kingdom network analyses between bacteria and fungi showed the complexity of the microbiome network was significantly reduced and the interactions between bacteria and fungi were also weakened in the CL-FMT group vs the CON-FMT group (Figure 4f). The above results showed that the trends of the gut microbiome in recipient mice were consistent with those in the donor mice. Then, the Venn diagram was used to assess the gut microbiome profile of FMT mice at the genus level. The results showed that 85.11% (40/47) of bacterial genera and 52.38% (33/63) of fungi genera present in the CL inocula were successfully transferred to the CL-recipient mice, and 91.45% (107/117) of bacteria genera and 56.36% (31/55) of fungi genera in the CON inocula were also successfully transferred to the CON-recipient mice, respectively (Figure 4g). These results indicated that the FMT model was successfully established, and the CL-recipient mice showed similar characteristics of gut microbiota dysbiosis with the CL group.

Gut microbiota dysbiosis enhances the susceptibility of MS pulmonary colonization in mice.

(a) The experimental procedures of the feces microbiota transplantation. (b) The PCoA analysis of gut microbiota based on weighted Unifrac distance. (c) The ITS2/16S diversity ratio at the genus level. (d) The relative abundant ratio of Firmicutes/Bacteroidota. (e) The relative abundant ratio of Ascomycota/Basidiomycota. (f) The trans-kingdom correlation networks of CON-FMT and CL-FMT groups at the genus level. (g) The Venn diagram of gut bacteria and gut fungi in different groups at the genus level. (h) The load of MS in the lungs of mice after infection at 72 h. FMT: fecal microbiota transplantation; CON-FMT: The fecal microbiota of the control group was transplanted, CL-FMT: The fecal microbiota of the clindamycin-treatment group was transplanted. MS: Mycobacterium smegmatis, * P<0.05.

Then, we assessed the susceptibility of MS in the lungs of FMT mice, the results showed that the burden of MS in the lungs of CL-recipient mice was significantly rise than that in the CON-recipient mice after 72 hours of MS infection (Figure 4h). However, the size of the cecum showed no significant differences between the two groups (Figure S4a). The level of iFABP and LPS also significantly increased in the CL-recipient mice compared with the CON-recipient mice after FMT (Figure S4b). The pathological sections of intestinal tissue showed the intestinal epithelial tissues of mice have no significant damage after FMT (Figure S4c). Altogether, our data suggested that gut microbiota dysbiosis increases intestinal permeability and enhances the susceptibility of MS colonization in the lungs of mice.

The gut microbiota dysbiosis altered the lung transcriptome and increased Nos2 expression

To further explore the potential mechanisms by which intestinal microbiota dysbiosis affects MS colonization in the lungs of mice, we performed a transcriptome analysis of the mice’s lung tissue. The results showed that there were 1191 up-regulated differentially expressed genes (DEGs) and 1013 down-regulated DEGs in the CL groups vs the CON group (FDR < 0.05, and FC >1, Figure S5a). Compared with the CON-FMT group, 274 DEGs were up-regulated and 32 DEGs were down-regulated in the CL-FMT group (Figure S5b). Then, we screened the overlapping DEGs between the two comparison sets including CL vs CON and CL-MFT vs CL-FMT. The Venn diagram showed that 93 upregulating-DEGs and 5 downregulating-DEGs were commonly expressed among these groups (Figure 5a). Subsequently, GO and KEGG pathway enrichment analyses were performed to clarify the function of these 98 DEGs. The top 30 biological processes enriched by GO showed that CL and CL-FMT groups mainly affected the immune response and inflammatory response, including “defense response”, “response to bacterium”, “cellular response to interleukin-1”, and “cellular response to lipopolysaccharide” (Figure 5b). The enriched molecular functions of these DEGs were some cytokine and protease involving the immune defense response, including “cytokine activity”, “chemokine activity”, “nitric-oxide synthase binding”, and “CXCR chemokine receptor binding” (Figure 5b).

Gut microbiota dysbiosis altered the lung transcriptome of mice.

(a) The Venn diagram of DEGs in the different groups. (b) The top 30 GO enrichment analyses of DEGs. (c) The top 20 KEGG enrichment analyses of DEGs. (d) Interaction network analysis of selected DEGs and significant KEGG pathways related to human diseases. (e) The expression levels of Nos2, Ctsd, Cd74, and Tnf with RT-qPCR and RNA-seq. DEGs: differentially expressed genes, CL: clindamycin-treatment group, CON: control group; CON-FMT: The fecal microbiota of the control group was transplanted, CL-FMT: The fecal microbiota of the clindamycin-treatment group was transplanted. * P<0.05, **P<0.01.

The top 20 KEGG pathways enrichment analyses at classification level 1 indicated that these DEGs are mainly involved in the process of environmental information processing, human diseases, and organismal systems (Figure 5c and Table S5). At classification level 2 of the KEGG analysis, we found that there were 6 pathways relating to infectious disease, and 7 pathways involving the immune system among the enrichment top 20 pathways (Figure 5c and Table S6). Interestingly, among the 6 pathways of infectious disease, there were 3 pathways involving bacterial infectious disease, including TB, legionellosis, and pertussis (Figure 5c and Table S6). In addition, the pathway of “Graft-versus-host disease” is also enriched in these DEGs (Figure 5c).

For the future screening of the potential regulatory genes correlated with TB, 9 KEGG pathways related to human disease were selected for interaction network analysis with DEGs. The results showed that 8 genes were strongly associated with TB (Figure 5d). Among those genes, 7 genes including Nos2, Cd14, Tnf, Cd74, Clec4e, Ctsd, and Il6 are up-regulating expression and the Cd209a gene is the down-regulating expression in the CL and CL-FMT groups (Figure S5). It is worth noting that the Nos2 gene was an enriched expression in both GO and KEGG analyses. Then, we performed an RT-qPCR to assess the validity of the transcriptome. The results revealed that the genes of Nos2, Ctsd, Cd74, and Tnf were increased in CL-FMT and CL groups compared with that in control groups (Figure 5e), which indicated that the RNA-Seq data is reliable. In summary, our data revealed that gut microbiota dysbiosis significantly altered the transcriptomic profiling in the lung tissue of mice, and increased the expression of Nos2 genes.

Nos2 regulates the expression of NO, ROS, and Defb1

Based on the above transcriptomic results, we speculate that the upregulating expression of Nos2 induced by gut microbiota dysbiosis may play an important role in MS pulmonary colonization in mice. Hence, an over-expression Nos2 vector, Nos2-pcDNA3.1, was constructed and transfected into A549 cells. Then the cells were infected with MS to verify the effect of Nos2 overexpression on MS colonization ability. The results showed that the expression level of Nos2 was significantly raised in A549 cells after transfection with the Nos2-pcDNA3.1 plasmid (Figure 6a). MS colonization density was significantly increased in A549 cells with Nos2-pcDNA3.1 plasmid after infection MS 6h, 12h, and 24h (Figure 6b), which suggested that Nos2 overexpression increases the infection susceptibility of MS to A549 cells.

Nos2 regulates the expression of NO, ROS, and Defb1.

(a) The expression levels of the Nos2 in the A549 cells. (b) The load of MS in the A549 cells at different times. (c) The NO concentration in the A549 cells. (d) The concentration of ROS in the A549 cells. (e) The expression levels of Defb1 in A549. A: A549 cells control, AP: Transfected A549 cells with blank pcDNA3.1plasmid, AN: Transfected A549 cells with Nos2-pcDNA3.1 plasmid. * P<0.05, **P<0.01, *** P<0.001.

Subsequently, we explored the potential molecular mechanism by which Nos2 over-expression increased MS colonization ability. Nos2 is a key enzyme required for nitric oxide (NO) synthesis, so we detected the level of NO in A549 cells. The results showed that the NO level significantly increased in the A549 cells with transfection of the Nos2-pcDNA3.1 plasmid (Figure 6c). Reactive oxygen species (ROS) play an important role in pathogens to invade and colonize in host organs15. To explore whether Nos2 is involved in regulating ROS production in A549 cells, ROS concentration in cells was detected by the fluorescent probe DCFH-DA. As fluorescence microscopy showed in Figure 6d, ROS concentration in A549 cells transfected with Nos2-pcDNA3.1 was significantly lower than that in the control A549 cells and A549 cells transfected with pcDNA3.1 empty vector (Figure 6d). In addition, we also detected the expression of human β-defensin-1 (Defb1), an important antimicrobial peptide16, by RT-qPCR. The results showed that Nos2 overexpression significantly reduced the level of Defb1 in A549 cells (Figure 6e). Taken together, our results revealed that Nos2 overexpression disrupts the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1, and results in the enhancement of MS pulmonary colonization capacity in mice.

Discussion

TB still remains a major health challenge globally. How to effectively limit or reduce Mtb colonization in the host is a potential strategy to prevent this disease transmission and development. Recent studies have shown that the gut microbiome can affect distant organs via the “gut-lung axis”, and gut microbiota dysbiosis is a potential factor inducing respiratory diseases including TB17,18. Previous studies have shown that the gut microbiota was significantly different in TB patients compared with healthy humans and revealed gut microbiota dysbiosis, especially Bacteroidota, and Firmicutes, is strongly associated with the development of TB8,9. However, how the gut microbiota affects TB development by the “gut-lung axis” remains unclear. In the present study, we established a mouse model of gut microbiota dysbiosis and an FMT model using clindamycin treatment and revealed the potential mechanisms of intestinal microbiota dysbiosis promoting MS pulmonary colonization in the mouse by up-regulating Nos2 gene-associated pathways.

Antibiotics can induce intestinal microbiota dysbiosis, which in turn affects host immunity and leads to increased susceptibility and deterioration of a variety of diseases19. Previous research has verified that gut Bacteroidota dysbiosis is strongly associated with the development of TB20–22. Hence, we select clindamycin, an antibiotic that selectively disrupts anaerobic Bacteroidota14, to treat the mice in this study. Our results show that the abundance of Bacteroidota was significantly reduced and the abundance of Firmicutes was increased in the CL group compared with the CON group. This result is consistent with the previous study which reported that CL is effective in clearing Bacteroidota14. Bacteroidota is a consortium of many commensal bacterial species responsible for major fermentation processes, glycolipid production, and promoting systemic Th1 immune responses23.

The fungi are also important components of the gut microbiome. The gut fungi mycobiota dysbiosis is related to many diseases24. In the present study, we found that the gut fungi mycobiota balance of mice was disrupted by CL, especially, Aspergillus and Cladosporium significant increase in the CL group vs the CON group. Aspergillus and Cladosporium are opportunistic pathogens that usually cause lung infections in immunocompromised patients25,26. Bacteria and fungi harbor together in the gastrointestinal tract and occupy the same ecological niche. They interact with each other and develop complex ecological networks through these interactions. Homeostasis of microbial ecological networks limits pathogen invasion and infection27. The ratio of ITS2/16S rRNA can reflect the diversity and composition structural alterations of the fungi-bacterial microbiota28. In general, antibiotics are directed against bacteria and do not affect fungi. However, our data revealed that CL treatment significantly altered the diversity and composition structure of the fungi-bacterial microbiota, and reduced the trans-kingdom network complexity and interaction between bacteria and fungi. These suggested that the alterations of gut bacteria can affect the composition and diversity of fungi. Previous documents had also reported that any changes in gut bacteria would inadvertently cause alterations in the gut fungal community, and targeted fungal interventions would also cause changes in the gut bacterial community29–32. Hence, Gut microbiota should be taken as a whole rather than a single genus to explain the relationship between gut microbiota and associated diseases.

The FMT results revealed that gut commensal bacteria dysbiosis can promote MS colonization in the lungs of mice, but the underlying mechanisms remain not fully understood. Recent studies have suggested that changes in the “gut-lung axis” are a contributing factor in the pathogenesis and clinical manifestations of Mtb infection33,34. What is the substance mediating the interaction between the lung and the gut? There have been a variety of results in the different studies. Yang and colleagues have found that the gut microbiota via modulation of lncRNA mediated protective immunity against Mtb, and they found Bacteroides fragilis directly upregulated expression of lncRNA and promoted anti-TB immunity33. However, another clinical study on TB found that the expression level of iNOS in the plasma of new-onset pulmonary TB patients was significantly higher than that of healthy humans35. In the present study, we found that Nos2 expression significantly rises in the CL group and CL-FMT group. Hence, we speculated that the gut microbiota may increase MS colonization in mouse lungs by upregulation of Nos2 gene expression and associated pathways. Lung epithelial cells, as first responders during Mtb infection, have been shown to play an important role in TB pathogenesis36,37. Our results also confirmed that overexpression of Nos2 in A549 cells (a human alveolar epithelial cell) can enhance MS colonization in these cells.

Nos2, also termed iNOS, is a homodimeric enzyme that is induced by immune stimulation in an independent of intracellular Ca2+ manner and plays an important role in infection, inflammation, immune regulation, and the control of intracellular bacterial pathogen infection38. Compared with Nos1 and Nos3, Nos2 has the highest efficiency for production NO39. NO has direct antibacterial effects and immune-modulating function to intracellular pathogens and was considered a key molecule in controlling pathogens infections38. It has been verified that reactive nitrogen intermediates produced by murine cells could inhibit Mtb infection40. The use of Nos2 knock-out mice and NO synthase inhibitors is impaired in the control of Mtb growth41,42. However, our data found that the high concentration of NO increased the colonization of MS in A549 cells. This result seems to contradict that NO has antibacterial activity. Some studies have also shown that NO can promote bacterial growth. Ivan Gusarov et al reported that NO increases the resistance of bacteria to a broad spectrum of antibiotics, enabling the bacteria to survive and share habitats with antibiotic-producing microorganisms43. Cole et al showed that host cell-derived NO promoted the escape of listerolysin-dependent bacteria from phagocytic vacuoles into the cytoplasm by inhibiting the proton-pumping activity of V-ATPase and delaying phagolysosome fusion44,45. Another study has shown that the maximal level of NO produced by human macrophages was not bactericidal or bacteriostatic to Mtb, and the number of viable mycobacteria was increased in macrophages that produced NO, and this is correlated with the expression of nitrate reductase46. Therefore, we speculate that intracellular NO synthesized by Nos2 has dual roles, including bactericidal at high concentrations and promoting bacterial growth at low concentrations.

ROS are commonly present in various habitats occupied by living organisms and the production of ROS appears as a very ancient host strategy for coping with pathogens 47,48. ROS has shown potent antimicrobial activity against many pathogens including bacteria, fungi, and viruses. The accumulation of ROS is required for killing or inhibiting bacteria49. Nitric oxide interferes with antimicrobial killing by suppressing ROS accumulation43. Hence, we assess the level of ROS in the over-expression of Nos2 plasmids A549 cells. The results show that the ROS concentration in A549 cells transfected with Nos2-pcDNA3.1 was significantly lower than that in the control A549 cells and A549 cells transfected with pcDNA3.1 empty vector. This suggested that the promotion of bacterial proliferation by Nos2 is dependent on lower ROS levels. The crosstalk between Nos2 and ROS has been investigated in other research. Zheng and colleagues reported that the lower ROS level induced by Nos2a significantly inhibited E. piscicida proliferation and infection in zebrafish50. Another study has also shown that Nos2-produced ROS have an important role in maintaining homeostasis of the gut microbiota and defense against bacterial translocation51. Therefore, regulating the Nos2-ROS pathway may enhance the host’s defense ability against pathogens.

β-defensin-1 is an antimicrobial peptide, which is mainly produced by various epithelial cells and is an important part of the innate immune response52. Defb1 is also expressed in the respiratory epithelium cells and protects the airways from respiratory pathogens53. Previous research has found that Defb1 has 98% killing activity against active Mtb H37Rv54, and also has an effective antibacterial effect against dormant mycobacteria55. Consistent with the above results, our data revealed that there was a strong negative correlation between Defb1 gene expression and MS colonization in A549 cells. The present study was mainly performed at the gene level and lacked verification at the protein level. We will verify the role of the Nos2-Defb1 pathway on MS colonization at the protein levels in future studies.

Altogether, the crosstalk between the gut microbiome and the lungs through the “gut-lung axis” is complex and may involve multiple mechanisms, including immune response, metabolic disorders, cytokines, and inflammation et al. However, In the present study, we found that gut microbiome dysbiosis induced by clindamycin disturbs the gut equilibrium between bacteria and fungi, alters the lung transcriptome and increases Nos2 expression. Then, the microbiota dysbiosis could enhance the pulmonary colonization of MS in mice by regulating the Nos2-NO, Nos2-ROS, and Nos2-Defb1 pathways (Figure 7). In the future, we need more experiments to explore the relationship between the intestinal microbiome and the Nos2 associated pathways in the lung, and to further explore new targets for the prevention and treatment of TB.

Mechanisms of the intestinal microbiome dysbiosis effect on the colonization of MS in the mouse lungs.

Gut microbiome dysbiosis and gut permeability increasing disrupt the lung transcriptome, and increase Nos2 expression through the “gut-lung axis”. Nos2 high expression weakens the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1 in the cells, and promotes MS colonization in the lungs of mice. MS: Mycobacterium smegmatis.

Conclusions

In conclusion, the present study reveals that (i) clindamycin treatment induces gut microbiome dysbiosis, disturbs the gut equilibrium between bacteria and fungi, reduces the interactions among bacterial-fungal trans-kingdom, and increases intestinal permeability. (ii) the intestinal microbiome dysbiosis alters the lung transcriptome, increases Nos2 expression, and enhances MS colonization in the lungs of mice. (iii) intestinal microbiota could promote the pulmonary colonization of MS in mice by regulating the expression of NO, ROS, and Defb1 through Nos2 associated pathways, and changing the intracellular antimicrobial and anti-inflammatory environment. Hence, regulating the gut microbiome balance is a potential strategy for reducing the risks of Mtb infections and transmission.

Materials and methods

Mouse husbandry and antibiotic treatment

Specific pathogen-free (SPF) C57BL/6 mice (6 ∼ 8 weeks) were purchased from SiBeiFu (Beijing) Biotechnology Co., Ltd (No. SCXK (Beijing) 2019-0010). The mice were reared in an animal laboratory with a temperature of 22±2°C and a relative humidity of 50±5%, for one week before the model started. After adaptive feeding, mice were randomly divided into two groups, including a control group (CON) and a clindamycin group (CL), each group had 10 mice. During the experimental procedure, the CON group was fed with PBS. Whereas the CL group received 10mg/kg clindamycin by oral gavage once a day for 14 days. Each oral gavage treatment did not exceed 200μl, and treatment was stopped 2 days before the mycobacteria infection. All experiments were conducted according to the Declaration of Helsinki and were approved by the Animal Care and Use Ethics Committee of Xinxiang Medical University (No. EC-023-098).

Bacterial strains and infection

Due to the strong infectivity of Mtb, Mycobacterium smegmatis (MS), a model strain for the study of TB, was used in the present experiments. MS was grown in Middlebrook 7H9 liquid medium supplemented with Glycerin 0.5% and Tween 80 0.05%. After 18-24 h of cultivation in a shaking incubator at 37 °C, MS were centrifuged at 5 000 rpm for 5 min, and the resulting pellet was suspended in sterile PBS to the concentration of 2ⅹ107 CFU per mL. Figure 1a shows the experiment design procedure. For infection, mice were anesthetized by injection of 4% chloral hydrate (7 μl/g) and infected intranasally with 107 CFU of MS. The mice were euthanized after infection 72 hours, and the left lung of each mouse was extracted and homogenized in PBS with 0.1% Tween 20. 10-fold serial dilutions were made in PBS with 0.1% Tween 20 and plated on Middlebrook 7H10 Agar plates, and colonies of MS were counted after 5 days of incubation at 37°C.

Fecal sample DNA extraction, 16S rRNA, and ITS sequencing

Fresh stool samples from each mouse were collected aseptically and the total genomic DNA from fecal samples was extracted using the Quick-DNA Kit for feces (Qiagen, Germany) according to the manufacturer’s instructions. The quality and concentration of DNA were determined using NanoDrop® ND-2000 spectrophotometer (Thermo Scientific Inc, USA) and 1.0% agarose gel electrophoresis. Each DNA sample amplifies the hypervariable V3-V4 regions of the bacterial 16S rRNA genes using the primers 338F (5′-ACTCCTACGGAGGCAGCAGCAGCA-3′) and 806R (5′-GGACTACHVGGTWTAAT-3′). The ITS2 regions of the fungi were amplified using primers ITS3F (5’-GCATCGATGAAGAACGCAGC-3’) and ITS4R (5’-TCCTCCGCTTATTGATATGC-3’). The PCR products were purified with the AxyPrep DNA Gel Extraction Kit (Axygen Biosciences, Union City, CA, USA) and quantified using Quantus™ Fluorometer (Promega, USA). Purified amplicons were pooled in equimolar amounts, and paired-end sequenced on an Illumina MiSeq PE300 platform (Illumina, San Diego, USA) according to the standard protocols by Majorbio Bio-Pharm Technology Co. Ltd. (Shanghai, China).

Serum iFABP and LPS Measurement

The blood samples of all mice were collected via Eyeball, and the serum was separated by centrifugation at 1, 000×g for 20 minutes. Then, ELISA technology was used to measure the iFABP and LPS, according to the manufacturer’s instructions. The commercial ELISA kits were obtained from Shanghai Enzyme-linked Biotechnology Co., Ltd (Shanghai, China) including iFABP (catalog number TW9968) and LPS (catalog number TW12543).

Fecal microbial transplantation

To establish the gut microbiota depletion mouse model, SPF mice were treated with a cocktail of antibiotics (1 mg/mL ampicillin, 1 mg/mL metronidazole, 1 mg/mL neomycin, and 0.5 mg/mL vancomycin) by oral gavage daily for 14 days. All mice were randomly separated into two groups, the CON-recipient group (n = 10) and the CL-recipient group (n = 10). Each group of mice was randomly housed in two cages with five mice in each cage. The fecal microbial transplantation procedure is shown in Figure 4a. Fecal samples from mice of the CON group and CL group were collected in sterile containers. Then 1 g fecal sample was suspended in 5 ml sterile PBS, followed by the vortexes, sedimentation, and filtrate with a 100-μm cell strainer. The suspension of the same group was mixed as microbiota donors and immediately administered to the mice by oral gavage. 200 μl of the supernatant containing fecal microbiota from either CON or CL donors was transferred to microbiota-depleted mice by oral gavage every day, for 14 days. All recipient mice were infected intranasally with 107 CFU MS after stopping FMT 2 days, and 3 days later of the infection, all mice were sacrificed, and the colonic contents and lung tissue samples were collected for further analysis.

Lung tissue RNA extraction and RNA sequencing

The lung tissue from each mouse was separated aseptically and quickly stored in liquid nitrogen for subsequent RNA extraction. Total RNA was extracted from lung tissue using TRIzol® reagent (Dingguo Changsheng Biotechnology Co., Ltd, Beijing, China) according to the manufacturer’s protocol. RNA quantity and quality were determined using the NanoDrop 2000 Spectrophotometer (Thermo Scientific, USA). The Agilent 2100 Bioanalyzer (Agilent Technologies, Santa Clara, CA, USA) was used to assess RNA integrity. RNA-seq libraries were prepared using VAHTS Universal V6 RNA-seq Library Prep Kit following the manufacturer’s recommendations. Then the paired-end RNA-seq libraries were sequenced with the Illumina HiSeq X Ten platform (2 × 150 bp read length) by OE Biotech Co., Ltd (Shanghai, China). After the quality control, clean reads performed bioinformatics analysis.

Quantification of gene expression using RT-qPCR

Total RNA was extracted from mouse lungs or A549 cells using the Trigol (Dingguo Changsheng Biotechnology Co., Ltd, Beijing, China) according to the manufacturer’s protocol, and was reversed transcribed using the kit from All-in-One Script RTpremix (Kryptoner Mei Co., Ltd, Zhengzhou, China). Then RT-qPCR was performed with TB Green Premix Ex Taq Ⅱ (TaKaRa Biotechnology, China) to evaluate the amount of mRNA expression according to the manufacturer’s recommendations. Subsequently, PCR products were detected on a sequence detection system. The primer sequences of RT-qPCR were listed in Table S7 in this study. The relative gene expression levels were calculated using the 2 −ΔΔCt method.

Construction of Nos2 over-expression Plasmid and transfection into A549 cells

To enhance Nos2 expression levels in A549 cells, the recombinant plasmid Nos2-pcDNA3.1 was constructed. Based on the nucleotide sequences of Nos2 (GenBank accession No: NM_010927), the forward and reverse primers Nos2F/Nos2R (Table S7) were used for cloning the open reading frame and inserted into pcDNA3.1 expression vector. The insert orientation was confirmed by separate XhoI and BamHI digests followed by agarose gel electrophoresis. The Nos2-pcDNA3.1 plasmid was mixed with lipo 8000 (Beyotime, Shanghai, China) and then transfected into A549 cells via electroporation. 24 hours post-transfection, transfection efficiency was determined using RT-qPCR to ensure Nos2 over-expression in A549 cells.

Mycobacterium smegmatis infects in A549 cells

A549, a human alveolar basal epithelial cell, was used to assess the infection capacity of MS. A549 Cells were inoculated into 24-well plates with a density of 2ⅹ105 cells per well, and were cultured in RPMI1640 medium (Gibco Laboratories, USA) supplemented with 10% fetal bovine serum (Sangon Biotech Co., Ltd., Shanghai, China) and 100 U/mL penicillin/streptomycin (Beyotime, Shanghai, China) at 37 °C. Nos2-pcDNA3.1 was transfected into A549 cells when the cell density reached 70-80%. After 24 h, the supernatant was replaced with fresh medium containing MS and cultured for 6h, 12 h, and 24h. Then the wells were washed 3 times with PBS to remove unattached bacteria. Subsequently, A549 cells were lysed with 1ml 0.1% TritonX-100 (Shanghai Beyotime Biotechnology Co., Ltd., Shanghai, China) per well to prepare appropriate dilutions, which were plated on MiddleBrook7H10 Agar plates cultured for standard colony counts.

Detection of NO, ROS, and Defb1

Generation of NO in the A549 cells was detected by Nitric Oxide (NO) assay kit (Nanjing Jiancheng Biotechnology Research Institute Co., Ltd, Nanjing, china). Briefly, Cells were seeded into 24-well plates at a density of 2ⅹ105 cells per well, and Nos2-pcDNA3.1 was transfected into A549 cells when the cell density reached 70-80%. After 24 h, the cells were harvested by trypsinization and centrifuged at 1,000 rpm for 10 minutes. Then NO concentration was detected with the spectrophotometric method according to the manufacturer’s protocols. ROS levels were measured using the DCFH-DA. Briefly, after 24 h post-transfection of Nos2-pcDNA3.1, DCFH-DA was added to the culture medium at a final concentration of 10 μmol/mL, and incubated for 1 h at 37°C. The cells were washed twice with PBS and then stained with 10 μmol/mL DCFH-DA at 25 °C for 30 min in the dark room. Images were acquired by a confocal microscope (Nikon, Japan). For ROS quantification, A549 cells were collected, rinsed twice with PBS, and suspended in 10 μmol/mL DCFH-DA for 30 min. After incubation, fluorescence was detected at 485 nm (excitation) and 530 nm (emission) using a microplate reader. All these analyses were conducted in three replicates. The expression of Defb1 was detected using RT-qPCR, and the primer sequences of RT-qPCR are listed in Table S7.

Bioinformatics analysis

For microbiome analysis, the bioinformatics data were analyzed using the Majorbio Cloud platform (https://cloud.majorbio.com). The alpha diversity at the genus level was assessed according to the Chao 1 and inverse Simpson index. The beta diversity was calculated by principal coordinate analysis (PCoA). A permutational analysis of variance was performed to assess the variation in the taxonomic structure of microbiota communities between groups. The Wilcoxon rank sum test was used to assess the different structures of microbiota communities between groups. LEfSe analyses were performed to compare different biomarkers between groups. The trans-kingdom network figures were built using the package igraph (version 1.2.6). For transcriptomic analysis, the bioinformatics data were analyzed using the OE Cloud platform (https://www.oebiotech.com). Differential expression analysis was performed using the DESeq2 q value < 0.05 and foldchange > 2 was set as the threshold for significantly differential expression genes (DEGs). Venn diagram and volcano plot of DEGs were performed to explore gene expression patterns. The functional enrichment of the above DEGs was conducted using the Gene Ontology database (GO) (http://www.geneontology.org/) and the Kyoto Encyclopedia of Genes and Genomes (KEGG) (http://www.genome.jp/kegg/).

Statistical analysis

All statistics were performed using GraphPad Prism 8.0. If the data followed a normal distribution, unpaired Student’s t-tests were used to compare various parameters between the two groups. If the data did not follow a normal distribution, a non-parametric Mann-Whitney U test, and Wilcoxon rank sum test were used to compare the results. One-way analysis of variance was used for three or more groups of data. The graphs were made with GraphPad Prism 8.0 or R package (version 3.6.2). P-values of <0.05 were set as a threshold for statistical significance.

eLife Assessment

This important study shows the effect of gut dysbiosis on the colonization of mycobacteria in the lung. The data with comprehensive analysis of gene expression profiles in the lung with dysbiotic mice is compelling and goes beyond the current state of the art. However, the mechanistic insight and the experiments with Mtb infection are incomplete. With those parts strengthened, this paper would be of interest to researchers working on Mtb infection.

Reviewer #1 (Public Review):

Summary:

This work sought to demonstrate that gut microbiota dysbiosis may promote the colonization of mycobacteria, and they tried to prove that Nos2 down-regulation was a key mediator of such gut-lung pathogenesis transition.

Strengths:

They did large-scale analysis of RNAs in lungs to analyze the gene expression of mice upon gut dysbiosis in MS-infected mice. This might help provide an overview of gene pathways and critical genes for lung pathology in gut dysbiosis. This data is somewhat useful and important for the TB field.

Weaknesses:

(1) They did not use wide-type Mtb strain (e.g. H37Rv) to develop mouse TB infection models, and this may lead to the failure of the establishment of TB granuloma and other TB pathology icons.

(2) The usage of in vitro assays based on A542 to examine the regulation function of Nos2 expression on NO and ROS may not be enough. A542 is not the primary Mtb infection target in the lungs.

(3) They did not examine the lung pathology upon gut dysbiosis to examine the true significance of increased colonization of Mtb.

(4) Most of the studies are based on MS-infected mouse models with a lack of clinical significance.

Reviewer #2 (Public Review):

The manuscript entitled "Intestinal microbiome dysbiosis increases Mycobacteria pulmonary colonization in mice by regulating the Nos2-associated pathways" by Han et al reported that using clindamycin, an antibiotic to selectively disorder anaerobic Bacteriodetes, intestinal microbiome dysbiosis resulted in Mycobacterium smegmatis (MS) colonization in the mice lungs. The authors found that clindamycin induced damage of the enterocytes and gut permeability and also enhanced the fermentation of cecum contents, which finally increased MS colonization in the mice's lungs. The study showed that gut microbiota dysbiosis up-regulated the Nos2 gene-associated pathways, leading to increased nitric oxide (NO) levels and decreased reactive oxygen species (ROS) and β-defensin 1 (Defb1) levels. These changes in the host's immune response created an antimicrobial and anti-inflammatory environment that favored MS colonization in the lungs. The findings suggest that gut microbiota dysbiosis can modulate the host's immune response and increase susceptibility to pulmonary infections by altering the expression of key genes and pathways involved in innate immunity. The authors reasonably provided experimental data and subsequent gene profiles to support their conclusion. Although the overall outcomes are convincing, there are several issues that need to be addressed:

(1) In Figure S1, the reviewer suggests checking the image sizes of the pathological sections of intestinal tissue from the control group and the CL-treatment group. When compared to the same intestinal tissue images in Figure S4, they do not appear to be consistently magnified at 40x. The numerical scale bars should be presented instead of just magnification such as "40x".

(2) In Figure 4d, the ratio of Firmicutes in the CL-FMT group decreased compared to the CON-FMT group, whereas the CL-treatment group showed an increase in Firmicutes compared to the Control group in Figure 3b. The author should explain this discrepancy and discuss its potential implications on the study's findings.

(3) In Figure 6, did the authors have a specific reason for selecting Nos2 but not Tnf for further investigation? The expression level of the Tnf gene appears to be the most significant in both RT-qPCR and RNA-sequencing results in Figure 5f. Tnf is an important cytokine involved in immune responses to bacterial infections, so it is also a factor that can influence NO, ROS, and Defb1 levels.

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1 (Public Review):

Summary:

This work sought to demonstrate that gut microbiota dysbiosis may promote the colonization of mycobacteria, and they tried to prove that Nos2 down-regulation was a key mediator of such gut-lung pathogenesis transition.

Strengths:

They did large-scale analysis of RNAs in lungs to analyze the gene expression of mice upon gut dysbiosis in MS-infected mice. This might help provide an overview of gene pathways and critical genes for lung pathology in gut dysbiosis. This data is somewhat useful and important for the TB field.

Weaknesses:

(1)They did not use wide-type Mtb strain (e.g. H37Rv) to develop mouse TB infection models, and this may lead to the failure of the establishment of TB granuloma and other TB pathology icons.

The colonization of M.tb in the lungs and the amount of colonization are the first and primary conditions for the occurrence of TB. Our aim in this study is to explore the impact of gut microbiota dysbiosis on the colonization of M.tb in the lungs. However, due to the lack of necessary conditions for biosafety in our laboratory, some highly infectious bacteria (such as M.tb) are not allowed to be cultured, and establishing the M.tb infection animal model in our laboratory does not meet the requirements of biosafety. Hence, we used the model strain of M.tb, M.smegmatis (MS), and established the animal-infected model for exploring the effect of gut microbiota dysbiosis on MS colonization in mice lungs. However, the establishment of MS infected model may not necessarily produce typical TB granulomas and other TB pathology signs. we have discussed the limitations of the current study in the discussion part of the manuscript. The suggested revisions are shown in lines 21-39 of page 15. In future studies, we plan to adopt the reviewer's suggestion and will use a wide-type M.tb strain to establish the TB-infected model in the laboratory that has biosafety standards to further verify the results of the current study.

(2) The usage of in vitro assays based on A542 to examine the regulation function of Nos2 expression on NO and ROS may not be enough. A542 is not the primary Mtb infection target in the lungs.

Thanks for the reviewer’s comments. Although alveolar epithelial cells (AECs) are not the main target cells of Mtb infection, they are among the cells that are contacted early in M.tb infection. Early M.tb invasion of AECs is very essential for the establishment of infection ( PMID 11479618). AECs are usually the initial site of the lung’s response against M.tb. Available literature suggests that freshly isolated AECs are more permissive to M.tb growth than macrophages（PMID 33228849）. As a cellular reservoir for M.tb, AECs are capable of facilitating rapid bacterial growth while potentially escaping recognition by phagocytes in the alveolus. The immune cells such as macrophages are the primary targets of M.tb infection, where the M.tb survive and proliferate, leading to the formation and maintenance of granulomas. However, AECs are subjected to the same density of infection, and the bacteria invade and replicate in these cells and induce cell apoptosis and necrosis, which is considered a major mechanism implicated in extra-pulmonary dissemination (PMID 12925134, PMID 32849525). Besides their direct barrier role, AECs also directly respond to M.tb infection by producing mediators such as cytokines, chemokines, and antimicrobial agents (PMID 35017314). Therefore, it is feasible to select alveolar epithelial cell A549 to explore the colonization mechanism of intestinal microbiota affecting M.tb in vitro.

(3) They did not examine the lung pathology upon gut dysbiosis to examine the true significance of increased colonization of Mtb.

We have added the results of the lung pathological section in the revised manuscript. The results of lung pathological sections are shown in lines 11-13 of page 4, and Figure S2 of supplement information.

(4) Most of the studies are based on MS-infected mouse models with a lack of clinical significance.

The first and primary condition of any pathogen infection is that the bacteria must invade the host through colonization and multiply in the target organ. This study aimed to investigate the effect of intestinal microbial dysbiosis on the colonization of mycobacterium in mouse lungs. Our laboratory does not meet the biosafety standard for culturing highly infectious bacteria such as Mycobacterium tuberculosis. So, we used the Mycobacterium smegmatis as a model strain for M.tb to establish the infected mice model in the current research. Although M. smegmatis is generally considered nonpathogenic, M. smegmatis is closely related to M.tb in biochemical characteristics, genetic information, cell structure, and metabolism( PMID 32674978). M.smegmatis is regarded as a valuable model organism in the study of M.tb, which is widely been used to explore the biological characteristics of M.tb such as physiological state, stress response, non-culture state reactivation, antimicrobial activity, and biochemical protection (PMID 32674978). It has also been reported that M.smegmatis could be used as a model strain to study the molecular mechanism of interaction between M.tb and its host (PMID 30546046, PMID25970481, PMID 29568875). However, in preclinical experimental research, we used M. smegmatis as the object of study. Instead of focusing on the pathological changes caused by M.smegmatis in the host lungs, we mainly focused on the influence of intestinal microbiota on the colonization of mycobacterium in the lungs and its possible mechanism, which provides a reliable model to study the prevention of early infection and spread of M.tb through regulating the intestinal microbiota. It has important clinical significance for the further development of new measures for the prevention and control of tuberculosis. If experimental conditions permit, the establishment of an infected model with wild-type M.tb can be used to verify the findings of the present study which may provide important clinical guidelines.

Reviewer #2 (Public Review):

The manuscript entitled "Intestinal microbiome dysbiosis increases Mycobacteria pulmonary colonization in mice by regulating the Nos2-associated pathways" by Han et al reported that using clindamycin, an antibiotic to selectively disorder anaerobic Bacteriodetes, intestinal microbiome dysbiosis resulted in Mycobacterium smegmatis (MS) colonization in the mice lungs. The authors found that clindamycin induced damage of the enterocytes and gut permeability and also enhanced the fermentation of cecum contents, which finally increased MS colonization in the mice's lungs. The study showed that gut microbiota dysbiosis up-regulated the Nos2 gene-associated pathways, leading to increased nitric oxide (NO) levels and decreased reactive oxygen species (ROS) and β-defensin 1 (Defb1) levels. These changes in the host's immune response created an antimicrobial and anti-inflammatory environment that favored MS colonization in the lungs. The findings suggest that gut microbiota dysbiosis can modulate the host's immune response and increase susceptibility to pulmonary infections by altering the expression of key genes and pathways involved in innate immunity. The authors reasonably provided experimental data and subsequent gene profiles to support their conclusion. Although the overall outcomes are convincing, there are several issues that need to be addressed:

(1) In Figure S1, the reviewer suggests checking the image sizes of the pathological sections of intestinal tissue from the control group and the CL-treatment group. When compared to the same intestinal tissue images in Figure S4, they do not appear to be consistently magnified at 40x. The numerical scale bars should be presented instead of just magnification such as "40x".

Thanks for the precise comments. We have carefully checked the pathological section in Figure S1 and Figure S5 and added the numerical scale bars to the figure. The revised sections are added in the supplementary materials.

(2) In Figure 4d, the ratio of Firmicutes in the CL-FMT group decreased compared to the CON-FMT group, whereas the CL-treatment group showed an increase in Firmicutes compared to the Control group in Figure 3b. The author should explain this discrepancy and discuss its potential implications on the study's findings.

The success of fecal microbial transfer (FMT) is influenced by many factors, such as host intestinal microbiota, immunity, and genetic factors (PMID 37167953). During FMT procedure, all microbiota of the donor feces do not have the same colonization ability in the recipients. Some research has revealed that the colonization success rate of Bacteroidetes is higher than that of Firmicutes [PMID 24637796]. In this study, we noticed that the reason for the difference between Figure 4D and Figure 3B was that during FMT, the colonization of Firmicutes decreased in the Cl-FMT receptor after transplantation, while the colonization of Bacteroides increased, resulting in a decrease in the proportion of Firmicutes/ Bacteroides in the Cl-FMT group. However, we considered the gut microbiota as a whole in the present study. After FMT, we found that 85.11% of bacterial genera and 52.38% of fungi genera present in the CL inocula were successfully transferred to the CL-recipient mice, and 91.45% of bacteria genera and 56.36% of fungi genera in the CON inocula were also successfully transferred to the CON-recipient mice, respectively (Figure 4g). The trans-kingdom network analyses between bacteria and fungi showed that the trends of the gut microbiome in recipient mice were consistent with those in the donor mice. Therefore, the FMT model established in this study remains successful. For reviewer clarification, we have added explanations in the discussion part of the manuscript. See lines 8-29 of page 12 for details.

(3) In Figure 6, did the authors have a specific reason for selecting Nos2 but not Tnf for further investigation? The expression level of the Tnf gene appears to be the most significant in both RT-qPCR and RNA-sequencing results in Figure 5f. Tnf is an important cytokine involved in immune responses to bacterial infections, so it is also a factor that can influence NO, ROS, and Defb1 levels.

Thanks for the valuable reviewer’s comment. By analyzing the transcriptome data, we found that there were 8 genes strongly associated with TB infection in the KEGG pathway, including Nos2, Cd14, Tnf, Cd74, Clec4e, Ctsd, Cd209a, and Il6. Then, we performed KO pathway analysis and found that the Nos2 gene was strongly associated with multiple pathways including “cytokine activity ", "chemokine activity", and "nitric oxide synthase binding". Moreover, in a clinical study on tuberculosis, the expression level of Nos2 in the plasma of patients with newly diagnosed tuberculosis was significantly higher than that of healthy people, indicating that Nos2 is associated with the occurrence of tuberculosis (PMID 34847295). Therefore, we selected Nos2 as the main target gene in the current study to conduct the correlation pathway analysis. As an important cytokine involved in the immune response to bacterial infection, Tnf mentioned by the reviewers may also be a factor affecting the levels of NO, ROS, and Defb1, which provides a new idea for our future research.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

First, they need to use a true Mtb-infected mouse model to determine the relationship between gut dysbiosis and increased lung infection of Mtb.

Second, the mechanism by which nos2-mediated NO and ROS production need to be further analyzed in the real Mtb infection process (either in vivo or in vitro）.

Third, Lung pathology should be included in addressing the increased colonization of mycobacteria. Addressing these problems may help improve this work.

(1) Our laboratory does not meet the biosafety standard for culturing highly infectious bacteria such as Mycobacterium tuberculosis. So, we used the Mycobacterium smegmatis as a model strain for M.tb to establish the infected mice model in the current research. Although M. smegmatis is generally considered nonpathogenic. M. smegmatis is closely related to M.tb in biochemical characteristics, genetic information, cell structure, and metabolism( PMID 32674978). M.smegmatis is regarded as a valuable model organism in the study of M.tb, and has been widely used to explore the biological characteristics of M.tb such as physiological state, stress response, non-culture state reactivation, antimicrobial activity, and biochemical protection (PMID 32674978). It has also been reported that M.smegmatis was used as a model strain to study the molecular mechanism of interaction between M.tb and its host (PMID 30546046, PMID25970481, PMID 29568875). However, in preclinical experimental research, we mainly focused on the influence of intestinal microbiota on the colonization of mycobacterium in the lungs and its possible mechanism which provides a reliable model to study the prevention of early infection and spread of M.tb through regulating the intestinal microbiota.

(2) In the future, we will establish an infected model with wild-type M.tb to verify the mechanism by which nos2-mediated NO and ROS production and promote M.tb colonization.

(3) We have added the results in the lung pathological section in the revised manuscript. The results of lung pathological sections are shown in lines 11-13 of page 4, and Figure S2 of supplement information.

Version 2

Abstract

Increasing researches reveal gut microbiota was associated with the development of tuberculosis. How to prevent or reduce Mycobacterium tuberculosis colonization in the lungs is a key measure to prevent tuberculosis. However, the data on gut microbiota preventing Mycobacterium colonization in the lungs were scarce. Here, we established the clindamycin-inducing intestinal microbiome dysbiosis and fecal microbial transplantation models in mice to identify gut microbiota’s effect on Mycobacterium’s colonization in the mouse lungs and explore its potential mechanisms. The results showed that clindamycin treatment altered the diversity and composition of the intestinal bacterial and fungal microbiome, weakened the trans-kingdom network interactions between bacteria and fungi, and induced gut microbiome dysbiosis in the mice. Gut microbiota dysbiosis increases intestinal permeability and enhances the susceptibility of Mycobacterium colonization in the lungs of mice. The potential mechanisms were gut microbiota dysbiosis altered the lung transcriptome and increased Nos2 expression through the “gut-lung axis”. Nos2 high expression disrupts the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1 in the cells, and resulting in promoting Mycobacteria colonization in the lungs of mice. The present study raises a potential strategy for reducing the risks of Mycobacteria infections and transmission by regulating the gut microbiome balance.

Introduction

Tuberculosis (TB) is a chronic infectious disease caused by Mycobacterium tuberculosis (Mtb), with approximately 10 million people being infected each year and 1.6 million deaths in 20211,2. The epidemic of TB has brought heavy economic and social burdens to countries around the world. According to the World Health Organization (WHO) Global TB Report 2022, the global spending on essential tuberculosis services for TB diagnosis, treatment, and prevention has exceeded $6.4 billion 3. Besides, Drug-resistant forms of Mtb are currently on course to be the world’s deadliest pathogens, responsible for a quarter of deaths due to antimicrobial resistance1. The prevalence of drug-resistant Mtb is also a major threat to the control of TB. Hence, it has been an urgent task for researchers to develop more effective measures for the prevention and treatment of TB.

Recent advances in gut microbiota explorations have led to improved knowledge of the “gut-lung axis”. Some studies have shown that intestinal microbiota dysbiosis may affect the occurrence and development of respiratory diseases including asthma, chronic obstructive pulmonary disease, and respiratory pathogens infection et al by the “gut-lung axis”4–6. The documents about the relationship between gut microbiota and TB have increased in recent years, and increasing evidence supports the presence of intestinal microbiota dysbiosis in TB patients. A case-control study of pulmonary TB in pediatric patients showed that the abundance of Prevotellaceae and Enterococcaceae were increased, and the abundance of Oscillospiraceae and Bifidobacteriaceae were decreased in TB patients compared with healthy control subjects7. Maji and colleagues have found that the levels of Provetalla decreased and Bacteroides increased in TB patients8. Recently, our teams’ study in active TB patients showed the abundance of Bacteroidetes significantly altered compared with health subjects 9. Another study also showed the phylum of Bacteroidetes altered in the feces samples of patients with recurrent TB10. All these researches indicated that Mtb infection can lead to the alteration of intestinal microbiota composition.

However, the causal relations between Mtb infection and intestinal microbiota dysbiosis have not been fully elucidated. Some factors including alcohol, smoking, HIV infection, and diabetes have been proven to cause gut microbiota dysbiosis, and these also are important risk factors for TB11. A recent study has shown that the mice fed with a high-fat diet-induced gut microbiota dysbiosis increased the risk of developing active TB in mice12. Another study found Firmicutes/Bacteroidetes ratio increased in a murine model of type 2 diabetes induced by an energy-dense diet, and increased the susceptibility to Mtb infection13. These findings support a viewpoint that the changes in the gut microbiome are a contributing factor in Mtb infection pathogenesis and its clinical presentation. Many documents supposed that gut microbiota dysbiosis could modulate immunity and inflammation reaction at distal sites such as the lungs, reduce colonization resistance by external pathogens, and promote TB susceptibility11,14. Hence, modulation of the gut microbiota and balance of the gut-lung axis was a potential avenue for TB prevention and management.

Based on our previous studies and other reports, we found the effect of Mtb infection on the gut Bacteroidetes was most significant. Hence, in the present study, clindamycin (CL), an antibiotic that selectively disorders anaerobic Bacteroidetes15, was used to establish a mouse model of intestinal microbiota dysbiosis. We uncovered the important role of gut microbiome dysbiosis affecting Mycobacteria colonization in the mouse lungs using the technology of fecal microbiota transplantation (FMT). Furthermore, based on the “gut-lung axis” theory, we performed a transcriptomic analysis of the mice lungs after Mycobacteria infection. We aim to explore the potential role and mechanism of gut microbiome dysbiosis enhancing Mycobacteria colonization in the lungs of mice.

Results

Clindamycin increases Mycobacterium smegmatis colonization in the lungs of mice

According to the Figure 1a procedure, we established a clindamycin-treated mice model to assess the effects of clindamycin-associated gut dysbiosis on MS colonization in the lungs of mice. The results showed that the mice with clindamycin treated presented a significantly higher colonization of MS in the lungs than that of control mice (Figure 1b). The size of the cecum of CL-treated mice was markedly more dilated compared with the control groups (Figure 1c). To assess the effect of CL on gut mucosal damage and permeability, the intestinal fatty acid binding protein (i-FABP, a marker of enterocyte death) and lipopolysaccharides (LPS, an endotoxin, as a marker of gut permeability) in serum were investigated with enzyme-linked immunosorbent assay (ELISA) technology. The results showed that the levels of i-FABP and LPS were significantly higher in the CL group than in the control group (CON) (Figure 1d and 1e). Pathological sections of intestinal tissue showed the intestinal epithelial tissues of mice with no obvious alteration after clindamycin treatment vs the CON group (Figure S1). H&E staining of paraffin-embedded lung sections revealed that there was more diffused inflammation and inflammatory cell infiltration in the CL-treated mice compared with that of the CON group animals after infecting with MS (Figure S2). All these results indicated that CL treatment damages the enterocyte, increases gut permeability, promotes the fermentation of cecum contents, and increases MS colonization in the lungs of mice.

Clindamycin increases MS pulmonary colonization in mice.

(a) The experimental procedures of clindamycin-inducing gut microbiota dysbiosis model in mice. (b) The colonization of MS in the lungs of mice after infection at 72 h. (c) The length of cecum after clindamycin treatment. (d)The level of iFABP in serum. (e) LPS concentration in serum. CL: clindamycin-treatment group, CON: control group. MS: Mycobacterium smegmatis, iFABP: intestinal fatty acid binding protein, LPS: lipopolysaccharides. *P<0.05, **P<0.01, *** P<0.001.

Altered the diversity and composition of the gut microbiome in mice

To assess the effect of CL treatment on the gut microbiome of mice, we investigated the alteration of gut microbiota (bacterial) and mycobiota (fungi) by 16S rRNA and ITS2 amplicon sequencing, respectively. For the gut bacterial analysis, the α-diversity was significantly decreased in the CL group compared with the CON group (Figure 2a). The β-diversity based on the weighted UniFrac distance showed that the CL group samples were separated from the CON group (R=0.8488, P=0.001) (Figure 2b). It indicated that the intestinal bacterial microbial community structure was a significant difference between the CL-treated mice and the control mice. At the phylum level, we observed a remarkable decrease of Bacteroidota and a significant increase of Firmicutes and Proteobacteria in the CL group compared to the control group (Figure 2c, Table S1). At the genus level, the CL group had a significantly higher relative abundance of Bacteroides, Lactobacillus, Escherichia-Shigella, and Faecalibaculum, and a lower abundance of Dubosiella, Alloprevotella, Akkermansia, and Bifidobacterium compared with that in the CON group (Figure 2d, Table S2).

Clindamycin altered the diversity and composition of the gut microbiome in mice.

(a) Chao1 and Inverse Simpson index of gut microbiota. (b) PCoA analysis of gut microbiota based on weighted Unifrac distance. (c) The composition distribution of gut microbiome (bacteria and fungi) at the phylum level. (d) The LEfSe analysis of the differentially abundant gut microbiota between the CL group and CON group at the genera level (LDA>4, P<0.05). CL: clindamycin-treatment group, CON: control group. **P<0.01, **** P<0.0001.

For the fungal mycobiota analysis, the Chao 1 and inverse Simpson index of the CL group was significantly increased compared to that of the CON group (Figure 2a). The PCoA showed that the samples were significantly distinguished between two groups (R=0.6038, P=0.001) (Figure 2b), which suggested that the community of intestinal mycobiota in the two groups is different. At the phylum level, the fungal microbiota was dominated by the Ascomycota and Basidiomycota in two groups, and Ascomycota showed no obvious differences between the two groups. However, the relative abundance of Basidiomycota significantly increases in the CL group compared with the CON group (Figure 2c. Table S3). The LEfSe analysis at the genus level showed that the relative abundance of Aspergillus, Penicillium, Cladosporium, and Alternaria was enriched in the CL group vs the CON group (Figure 2d, Table S4). These results are consistent with the Wilcoxon rank-sum test analysis (Figure S3). Taken together, all these data indicated that CL treatment altered the diversity and composition of the intestinal microbiome (including bacterial and fungi) in the mice and induced gut microbiome dysbiosis.

Clindamycin weakens the trans-kingdom networks construction of gut bacterial and fungi

To evaluate the effect of clindamycin treatment on gut microbiome balance, we analyzed the fungi-to-bacteria species ratio with ITS/16S. The results show that the ITS/16S ratio of the CL group was significantly increased compared with the CON group (Figure 3a). We also assessed the alterations in the ratio of the dominant phyla in the gut microbiome. The results showed that the ratio of Firmicutes/Bacteroidota in the CL group was increased, and Ascomycota/Basidiomycota was significantly decreased compared with that of the CON group (Figure 3b and 3c). These results revealed that clindamycin treatment not only disturbed the intestinal equilibrium between bacteria and fungi but also destroyed the balance among the dominant bacteria (or fungi). Then, to assess the interplay between bacteria and fungi, we performed the trans-kingdom networks analysis at the genus level. The results showed that the trans-kingdom networks were altered in the CL group vs that in the CON group. In the control group, the bacteria and fungi were closely related to each other, gathering in a cluster and forming a more complex network (Figure 3d). There were 154 nodes including 45 fungi and 109 bacteria, and 810 edges; the relative connectedness was 5.26 (Table 1). However, the complexity of the trans-kingdom network in the CL group was dramatically decreased, and the interplay between bacteria and fungi was also weakened (Figure 3e). A total of 110 nodes including 62 fungi and 48 bacteria gathered into clusters in this network. The edges decreased to 357 and the relative connectedness was 3.25 in the CL group network (Table 1). We also found that the interplay ratio between fungi in the CL group was 52.10%, which was much higher than that of the CON group (7.53%) (Table 1). All these results indicated that clindamycin treatment not only altered the proportion of bacteria and fungi, but also weakened the trans-kingdom networks, changed the interactions between bacteria and fungi, and resulted in gut microbiome dysbiosis.

Clindamycin weakens the trans-kingdom network construction of gut bacteria and fungi.

(a) The ratio of ITS2/16S at the genus level. (b) The relative abundant ratio of Firmicutes/Bacteroidota. (c) The relative abundant ratio of Ascomycota/Basidiomycota. (d) The trans-kingdom correlation networks of the CON at the genus level. (e) The trans-kingdom correlation networks of the CL at the genus level. CL: clindamycin-treatment group, CON: control group. *P<0.05; ***P<0.001.

Fecal microbial transfer from Clindamycin-treated mice donors to antibiotics-treated conventional mice enhances MS colonization

We performed an FMT model to further verify the relationship of gut microbiota dysbiosis with the susceptibility to MS infection in the mice’s lungs. The experimental design for the FMT is shown in the Figure 4a. First, we analyzed the α-diversity of gut bacterial and fungal microbiota after FMT with the Chao 1 index and inverse Simpson index. The results show that the α-diversity of gut bacterial microbiota was significantly decreased, and the α-diversity of gut fungal microbiota was significantly increased in the CL-FMT group compared with that in the CON-FMT group (Figure S4). Then we performed PCoA analysis to distinguish the gut microbiome alteration between CL- and CON-recipient mice. The results show a significant separation of gut bacterial and fungal microbiota between CT- and AD-recipient samples (p=0.007 and p=0.042, Figure 4b). The fungal-to-bacterial species ratio was significantly increased in the CL-recipient group vs the CON-recipient group (p<0.05, Figure 4c). To calculate the changes of the dominant phyla in the gut microbiome, the results show that Firmicutes/Bacteroidota in the CL-FMT group was decreased, and Ascomycota/Basidiomycota was no obvious different compared with that of the CON-FMT group (Figure 4d and 4e). The trans-kingdom network analyses between bacteria and fungi showed the complexity of the microbiome network was significantly reduced and the interactions between bacteria and fungi were also weakened in the CL-FMT group vs the CON-FMT group (Figure 4f). The above results showed that the trends of the gut microbiome in recipient mice were consistent with those in the donor mice. Then, the Venn diagram was used to assess the gut microbiome profile of FMT mice at the genus level. The results showed that 85.11% (40/47) of bacterial genera and 52.38% (33/63) of fungi genera present in the CL inocula were successfully transferred to the CL-recipient mice, and 91.45% (107/117) of bacteria genera and 56.36% (31/55) of fungi genera in the CON inocula were also successfully transferred to the CON-recipient mice, respectively (Figure 4g). These results indicated that the FMT model was successfully established, and the CL-recipient mice showed similar characteristics of gut microbiota dysbiosis with the CL group.

Gut microbiota dysbiosis enhances the susceptibility of MS pulmonary colonization in mice.

(a) The experimental procedures of the feces microbiota transplantation. (b) The PCoA analysis of gut microbiota based on weighted Unifrac distance. (c) The ITS2/16S diversity ratio at the genus level. (d) The relative abundant ratio of Firmicutes/Bacteroidota. (e) The relative abundant ratio of Ascomycota/Basidiomycota. (f) The trans-kingdom correlation networks of CON-FMT and CL-FMT groups at the genus level. (g) The Venn diagram of gut bacteria and gut fungi in different groups at the genus level. (h) The load of MS in the lungs of mice after infection at 72 h. FMT: fecal microbiota transplantation; CON-FMT: The fecal microbiota of the control group was transplanted, CL-FMT: The fecal microbiota of the clindamycin-treatment group was transplanted. MS: Mycobacterium smegmatis, * P<0.05.

Then, we assessed the susceptibility of MS in the lungs of FMT mice, the results showed that the burden of MS in the lungs of CL-recipient mice was significantly rise than that in the CON-recipient mice after 72 hours of MS infection (Figure 4h). However, the size of the cecum showed no significant differences between the two groups (Figure S5a). The level of iFABP and LPS also significantly increased in the CL-recipient mice compared with the CON-recipient mice after FMT (Figure S5b). The pathological sections of intestinal tissue showed the intestinal epithelial tissues of mice have no significant damage after FMT (Figure S5c). Altogether, our data suggested that gut microbiota dysbiosis increases intestinal permeability and enhances the susceptibility of MS colonization in the lungs of mice.

The gut microbiota dysbiosis altered the lung transcriptome and increased Nos2 expression

To further explore the potential mechanisms by which intestinal microbiota dysbiosis affects MS colonization in the lungs of mice, we performed a transcriptome analysis of the mice’s lung tissue. The results showed that there were 1191 up-regulated differentially expressed genes (DEGs) and 1013 down-regulated DEGs in the CL groups vs the CON group (FDR < 0.05, and FC >1, Figure S6a). Compared with the CON-FMT group, 274 DEGs were up-regulated and 32 DEGs were down-regulated in the CL-FMT group (Figure S6b). Then, we screened the overlapping DEGs between the two comparison sets including CL vs CON and CL-MFT vs CL-FMT. The Venn diagram showed that 93 upregulating-DEGs and 5 downregulating-DEGs were commonly expressed among these groups (Figure 5a). Subsequently, GO and KEGG pathway enrichment analyses were performed to clarify the function of these 98 DEGs. The top 30 biological processes enriched by GO showed that CL and CL-FMT groups mainly affected the immune response and inflammatory response, including “defense response”, “response to bacterium”, “cellular response to interleukin-1”, and “cellular response to lipopolysaccharide” (Figure 5b). The enriched molecular functions of these DEGs were some cytokine and protease involving the immune defense response, including “cytokine activity”, “chemokine activity”, “nitric-oxide synthase binding”, and “CXCR chemokine receptor binding” (Figure 5b).

Gut microbiota dysbiosis altered the lung transcriptome of mice.

(a) The Venn diagram of DEGs in the different groups. (b) The top 30 GO enrichment analyses of DEGs. (c) The top 20 KEGG enrichment analyses of DEGs. (d) Interaction network analysis of selected DEGs and significant KEGG pathways related to human diseases. (e) The expression levels of Nos2, Ctsd, Cd74, and Tnf with RT-qPCR and RNA-seq. DEGs: differentially expressed genes, CL: clindamycin-treatment group, CON: control group; CON-FMT: The fecal microbiota of the control group was transplanted, CL-FMT: The fecal microbiota of the clindamycin-treatment group was transplanted. * P<0.05, **P<0.01.

The top 20 KEGG pathways enrichment analyses at classification level 1 indicated that these DEGs are mainly involved in environmental information processing, human diseases, and organismal systems (Figure 5c and Table S5). At classification level 2 of the KEGG analysis, we found 6 pathways relating to infectious disease, and 7 pathways involving the immune system among the enrichment top 20 pathways (Figure 5c and Table S6). Interestingly, among the 6 pathways of infectious disease, there were 3 pathways involving bacterial infectious disease, including TB, legionellosis, and pertussis (Figure 5c and Table S6). In addition, the pathway of “Graft-versus-host disease” is also enriched in these DEGs (Figure 5c).

For the future screening of the potential regulatory genes correlated with TB, 9 KEGG pathways related to human disease were selected for interaction network analysis with DEGs. The results showed that 8 genes were strongly associated with TB (Figure 5d). Among those genes, 7 genes including Nos2, Cd14, Tnf, Cd74, Clec4e, Ctsd, and Il6 are up-regulating expression and the Cd209a gene is the down-regulating expression in the CL and CL-FMT groups (Figure S7). It is worth noting that the Nos2 gene was an enriched expression in both GO and KEGG analyses. Then, we performed an RT-qPCR to assess the validity of the transcriptome. The results revealed that the genes of Nos2, Ctsd, Cd74, and Tnf were increased in CL-FMT and CL groups compared with that in control groups (Figure 5e), which indicated that the RNA-Seq data is reliable. In summary, our data revealed that gut microbiota dysbiosis significantly altered the transcriptomic profiling in the lung tissue of mice, and increased the expression of Nos2 genes.

Nos2 regulates the expression of NO, ROS, and Defb1

Based on the above transcriptomic results, we speculate that the upregulating expression of Nos2 induced by gut microbiota dysbiosis may play an important role in MS pulmonary colonization in mice. Hence, an over-expression Nos2 vector, Nos2-pcDNA3.1, was constructed and transfected into A549 cells. Then the cells were infected with MS to verify the effect of Nos2 overexpression on MS colonization ability. The results showed that the expression level of Nos2 was significantly raised in A549 cells after transfection with the Nos2-pcDNA3.1 plasmid (Figure 6a). MS colonization density was significantly increased in A549 cells with Nos2-pcDNA3.1 plasmid after infection MS 6h, 12h, and 24h (Figure 6b), which suggested that Nos2 overexpression increases the infection susceptibility of MS to A549 cells.

Nos2 regulates the expression of NO, ROS, and Defb1.

(a) The expression levels of the Nos2 in the A549 cells. (b) The load of MS in the A549 cells at different times. (c) The NO concentration in the A549 cells. (d) The concentration of ROS in the A549 cells. (e) The expression levels of Defb1 in A549. A: A549 cells control, AP: Transfected A549 cells with blank pcDNA3.1plasmid, AN: Transfected A549 cells with Nos2-pcDNA3.1 plasmid. * P<0.05, **P<0.01, *** P<0.001.

Subsequently, we explored the potential molecular mechanism by which Nos2 over-expression increased MS colonization ability. Nos2 is a key enzyme required for nitric oxide (NO) synthesis, so we detected the level of NO in A549 cells. The results showed that the NO level significantly increased in the A549 cells with transfection of the Nos2-pcDNA3.1 plasmid (Figure 6c). Reactive oxygen species (ROS) play an important role in pathogens invade and colonize host organs16. To explore whether Nos2 regulates ROS production in A549 cells, ROS concentration in cells was detected by the fluorescent probe DCFH-DA. As fluorescence microscopy showed in Figure 6d, ROS concentration in A549 cells transfected with Nos2-pcDNA3.1 was significantly lower than that in the control A549 cells and A549 cells transfected with pcDNA3.1 empty vector (Figure 6d). In addition, we also detected the expression of human β-defensin-1 (Defb1), an important antimicrobial peptide17, by RT-qPCR. The results showed that Nos2 overexpression significantly reduced the level of Defb1 in A549 cells (Figure 6e). Taken together, our results revealed that Nos2 overexpression disrupts the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1, and resulting in the enhancement of MS pulmonary colonization capacity in mice.

Discussion

TB remains a major health challenge globally. How to effectively limit or reduce Mtb colonization in the host is a potential strategy to prevent this disease transmission and development. Recent studies have shown that the gut microbiome can affect distant organs via the “gut-lung axis”, and gut microbiota dysbiosis is a potential factor inducing respiratory diseases including TB18,19. Previous studies have shown that the gut microbiota was significantly different in TB patients compared with healthy humans and revealed gut microbiota dysbiosis, especially Bacteroidota, and Firmicutes, is strongly associated with the development of TB8,10. However, how the gut microbiota affects TB development by the “gut-lung axis” remains unclear. In the present study, we established a mouse model of gut microbiota dysbiosis and an FMT model using clindamycin treatment and revealed the potential mechanisms of intestinal microbiota dysbiosis promoting MS pulmonary colonization in the mouse by up-regulating Nos2 gene-associated pathways.

Antibiotics can induce intestinal microbiota dysbiosis, which in turn affects host immunity and leads to increased susceptibility and deterioration of a variety of diseases20. Previous research has verified that gut Bacteroidota dysbiosis is strongly associated with the development of TB21–23. Hence, we select clindamycin, an antibiotic that selectively disrupts anaerobic Bacteroidota15, to treat the mice in this study. Our results show that the abundance of Bacteroidota was significantly reduced and the abundance of Firmicutes was increased in the CL group compared with the CON group. This result is consistent with the previous study which reported that CL is effective in clearing Bacteroidota15. Bacteroidota is a consortium of many commensal bacterial species responsible for major fermentation processes, glycolipid production, and promoting systemic Th1 immune responses24.

The fungi are also important components of the gut microbiome. The gut fungi mycobiota dysbiosis is related to many diseases25. In the present study, we found that the gut fungi mycobiota balance of mice was disrupted by CL, especially, Aspergillus and Cladosporium significant increase in the CL group vs the CON group. Aspergillus and Cladosporium are opportunistic pathogens that usually cause lung infections in immunocompromised patients26,27. Bacteria and fungi harbor together in the gastrointestinal tract and occupy the same ecological niche. They interact with each other and develop complex ecological networks through these interactions. Homeostasis of microbial ecological networks limits pathogen invasion and infection28. The ratio of ITS2/16S rRNA can reflect the diversity and composition structural alterations of the fungi-bacterial microbiota29. In general, antibiotics are directed against bacteria and do not affect fungi. However, our data revealed that CL treatment significantly altered the diversity and composition structure of the fungi-bacterial microbiota, and reduced the trans-kingdom network complexity and interaction between bacteria and fungi. These suggested that the alterations of gut bacteria can affect the composition and diversity of fungi. Previous documents had also reported that any changes in gut bacteria would inadvertently cause alterations in the gut fungal community, and targeted fungal interventions would also cause changes in the gut bacterial community30–33. Hence, Gut microbiota should be taken as a whole rather than a single genus to explain the relationship between gut microbiota and associated diseases.

Since most of the microbes in the gut are non-culturable, the FMT is considered to be an ideal model to study the function of gut microbiota. The success or not of FMT is influenced by many factors, such as host intestinal microbiota, immunity, and genetic factors34. During the FMT, not all microbiota in the donor feces have the same colonization ability in the receptors. Some research has revealed that the colonization success rate of Bacteroidetes is higher than that of Firmicutes 35. In the present study, we found that the ratio of Firmicutes/ Bacteroidetes in the CL-treatment group increased compared to the control group (Figure 3b). However, after FMT, the ratio of Firmicutes/ Bacteroidetes decreased in the CL-FMT group vs the CON-FMT group (Figure 4d). We speculated that the possible reason for this difference was that the colonization of Firmicutes decreased in the CL-FMT receptor group after transplantation. In contrast, the colonization of Bacteroides increased, resulting in a decrease in the proportion of Firmicutes/ Bacteroides in the CL-FMT group. However, we considered the gut microbiota as a whole in the analysis of the experimental results. After FMT, we found that 85.11% of bacterial genera and 52.38% of fungi genera present in the CL inocula were successfully transferred to the CL-recipient mice, and 91.45% of bacteria genera and 56.36% of fungi genera in the CON inocula were also successfully transferred to the CON-recipient mice, respectively (Figure 4g). The trans-kingdom network analyses between bacteria and fungi showed that the trends of the gut microbiome in recipient mice were consistent with those in the donor mice. Therefore, the inconsistent results between Figure 3b and Figure 4d will not affect this study’s findings, and the FMT model established in this study is successful.

The FMT results revealed that gut commensal bacteria dysbiosis can promote MS colonization in the lungs of mice, but the underlying mechanisms remain not fully understood. Recent studies have suggested that changes in the “gut-lung axis” are a contributing factor in the pathogenesis and clinical manifestations of Mtb infection 36,37. What is the substance mediating the interaction between the lung and the gut? There have been a variety of results in the different studies. Yang and colleagues have found that the gut microbiota via modulation of lncRNA mediated protective immunity against Mtb, and they found Bacteroides fragilis directly upregulated expression of lncRNA and promoted anti-TB immunity36. However, another clinical study on TB found that the expression level of iNOS in the plasma of new-onset pulmonary TB patients was significantly higher than that of healthy humans38. In the present study, we found that Nos2 expression significantly rises in the CL group and CL-FMT group. Hence, we speculated that the gut microbiota may increase MS colonization in mouse lungs by upregulation of Nos2 gene expression and associated pathways. Lung epithelial cells, as first responders during Mtb infection, have been shown to play an important role in TB pathogenesis39,40. Our results also confirmed that overexpression of Nos2 in A549 cells (a human alveolar epithelial cell) can enhance MS colonization in these cells.

Nos2, also termed iNOS, is a homodimeric enzyme that is induced by immune stimulation in an independent of intracellular Ca2+ manner and plays an important role in infection, inflammation, immune regulation, and the control of intracellular bacterial pathogen infection41. Compared with Nos1 and Nos3, Nos2 has the highest efficiency for production NO42. NO has direct antibacterial effects and immune-modulating function to intracellular pathogens and was considered a key molecule in controlling pathogens infections41. It has been verified that reactive nitrogen intermediates produced by murine cells could inhibit Mtb infection43. The use of Nos2 knock-out mice and NO synthase inhibitors is impaired in the control of Mtb growth44,45. However, our data found that the high concentration of NO increased the colonization of MS in A549 cells. This result seems to contradict that NO has antibacterial activity. Some studies have also shown that NO can promote bacterial growth. Ivan Gusarov et al reported that NO increases the resistance of bacteria to a broad spectrum of antibiotics, enabling the bacteria to survive and share habitats with antibiotic-producing microorganisms46. Cole et al showed that host cell-derived NO promoted the escape of listerolysin-dependent bacteria from phagocytic vacuoles into the cytoplasm by inhibiting the proton-pumping activity of V-ATPase and delaying phagolysosome fusion47,48. Another study has shown that the maximal level of NO produced by human macrophages was not bactericidal or bacteriostatic to Mtb, and the number of viable mycobacteria was increased in macrophages that produced NO, and this is correlated with the expression of nitrate reductase49. Therefore, we speculate that intracellular NO synthesized by Nos2 has dual roles, including bactericidal at high concentrations and promoting bacterial growth at low concentrations.

ROS are commonly present in various habitats occupied by living organisms and the production of ROS appears as a very ancient host strategy for coping with pathogens 50,51. ROS has shown potent antimicrobial activity against many pathogens including bacteria, fungi, and viruses. The accumulation of ROS is required for killing or inhibiting bacteria52. Nitric oxide interferes with antimicrobial killing by suppressing ROS accumulation46. Hence, we assess the level of ROS in the over-expression of Nos2 plasmids A549 cells. The results show that the ROS concentration in A549 cells transfected with Nos2-pcDNA3.1 was significantly lower than that in the control A549 cells and A549 cells transfected with pcDNA3.1 empty vector. This suggested that the promotion of bacterial proliferation by Nos2 is dependent on lower ROS levels. The crosstalk between Nos2 and ROS has been investigated in other research. Zheng and colleagues reported that the lower ROS level induced by Nos2a significantly inhibited E. piscicida proliferation and infection in zebrafish53. Another study has also shown that Nos2-produced ROS have an important role in maintaining homeostasis of the gut microbiota and defense against bacterial translocation54. Therefore, regulating the Nos2-ROS pathway may enhance the host’s defense ability against pathogens.

β-defensin-1 is an antimicrobial peptide, which is mainly produced by various epithelial cells and is an important part of the innate immune response55. Defb1 is also expressed in the respiratory epithelium cells and protects the airways from respiratory pathogens56. Previous research has found that Defb1 has 98% killing activity against active Mtb H37Rv57, and also has an effective antibacterial effect against dormant mycobacteria58. Consistent with the above results, our data revealed that there was a strong negative correlation between Defb1 gene expression and MS colonization in A549 cells. The present study was mainly performed at the gene level and lacked verification at the protein level. We will verify the role of the Nos2-Defb1 pathway on MS colonization at the protein levels in future studies.

Altogether, the crosstalk between the gut microbiome and the lungs through the “gut-lung axis” is complex and may involve multiple mechanisms, including immune response, metabolic disorders, cytokines, and inflammation et al. However, In the present study, we found that gut microbiome dysbiosis induced by clindamycin disturbs the gut equilibrium between bacteria and fungi, alters the lung transcriptome, and increases Nos2 expression. Then, the microbiota dysbiosis could enhance the pulmonary colonization of MS in mice by regulating the Nos2-NO, Nos2-ROS, and Nos2-Defb1 pathways (Figure 7). In the future, we need more experiments to explore the relationship between the intestinal microbiome and the Nos2-associated pathways in the lung and to further explore new targets for the prevention and treatment of TB.

Mechanisms of the intestinal microbiome dysbiosis effect on the colonization of MS in the mouse lungs.

Gut microbiome dysbiosis and gut permeability-increasing disrupt the lung transcriptome, and increase Nos2 expression through the “gut-lung axis”. Nos2 high expression weakens the intracellular antimicrobial and anti-inflammatory environment by increasing the concentration of NO, decreasing the levels of ROS and Defb1 in the cells, and promoting MS colonization in the lungs of mice. MS: Mycobacterium smegmatis.

Study limitations

This study has some limitations that need to be mentioned. First, due to the lack of experimental conditions in our laboratory that meet biosafety standards, we did not choose a wild-type Mtb strain to develop a mouse infection model, which may lead to the failure of the establishment of TB granuloma and other TB pathology icons and these also weaken the clinical significance of this study. Second, alveolar epithelial cells are one of the early contacting cells in Mtb infection, we only select alveolar epithelial cells (A549) to explore the colonization mechanism of intestinal microbiota affecting MS in vivo. This study lacks the choice of Mtb infection target cells-alveolar macrophages as the research object. Hence, future studies need to choose wild-type Mtb to establish animal infection models and alveolar macrophages for in vitro experiments to explore the regulation function of Nos2 expression on NO, ROS, and Defb1, which will resolve these limitations of the study.

Conclusions

In conclusion, the present study reveals that (i) clindamycin treatment induces gut microbiome dysbiosis, disturbs the gut equilibrium between bacteria and fungi, reduces the interactions among bacterial-fungal trans-kingdom, and increases intestinal permeability. (ii) the intestinal microbiome dysbiosis alters the lung transcriptome, increases Nos2 expression, and enhances MS colonization in the lungs of mice. (iii) intestinal microbiota could promote the pulmonary colonization of MS in mice by regulating the expression of NO, ROS, and Defb1 through Nos2-associated pathways, and changing the intracellular antimicrobial and anti-inflammatory environment. Hence, regulating the gut microbiome balance may be a potential strategy for reducing the risks of Mtb infections and transmission.

Materials and methods

Mouse husbandry and antibiotic treatment

Specific pathogen-free (SPF) C57BL/6 mice (6 ∼ 8 weeks) were purchased from SiBeiFu (Beijing) Biotechnology Co., Ltd (No. SCXK (Beijing) 2019-0010). The mice were reared in an animal laboratory with a temperature of 22±2°C and a relative humidity of 50±5%, for one week before the model started. After adaptive feeding, mice were randomly divided into two groups, including a control group (CON) and a clindamycin group (CL), each group had 10 mice. During the experimental procedure, the CON group was fed with PBS. Whereas the CL group received 10mg/kg clindamycin by oral gavage once a day for 14 days. Each oral gavage treatment did not exceed 200μl, and treatment was stopped 2 days before the mycobacteria infection. All experiments were conducted according to the Declaration of Helsinki and were approved by the Animal Care and Use Ethics Committee of Xinxiang Medical University (No. EC-023-098).

Bacterial strains and infection

Due to the strong infectivity of Mtb, Mycobacterium smegmatis (MS), a model strain for the study of TB, was used in the present experiments. MS was grown in Middlebrook 7H9 liquid medium supplemented with Glycerin 0.5% and Tween 80 0.05%. After 18-24 h of cultivation in a shaking incubator at 37 °C, MS were centrifuged at 5 000 rpm for 5 min, and the resulting pellet was suspended in sterile PBS to the concentration of 2[107 CFU per mL. Figure 1a shows the experiment design procedure. For infection, mice were anesthetized by injection of 4% chloral hydrate (7 μl/g) and infected intranasally with 107 CFU of MS. The mice were euthanized after infection 72 hours, and the left lung of each mouse was extracted and homogenized in PBS with 0.1% Tween 20. 10-fold serial dilutions were made in PBS with 0.1% Tween 20 and plated on Middlebrook 7H10 Agar plates, and colonies of MS were counted after 5 days of incubation at 37°C.

Fecal sample DNA extraction, 16S rRNA, and ITS sequencing

Fresh stool samples from each mouse were collected aseptically and the total genomic DNA from fecal samples was extracted using the Quick-DNA Kit for feces (Qiagen, Germany) according to the manufacturer’s instructions. The quality and concentration of DNA were determined using NanoDrop® ND-2000 spectrophotometer (Thermo Scientific Inc, USA) and 1.0% agarose gel electrophoresis. Each DNA sample amplifies the hypervariable V3-V4 regions of the bacterial 16S rRNA genes using the primers 338F (5′-ACTCCTACGGAGGCAGCAGCAGCA-3′) and 806R (5′-GGACTACHVGGTWTAAT-3′). The ITS2 regions of the fungi were amplified using primers ITS3F (5’-GCATCGATGAAGAACGCAGC-3’) and ITS4R (5’-TCCTCCGCTTATTGATATGC-3’). The PCR products were purified with the AxyPrep DNA Gel Extraction Kit (Axygen Biosciences, Union City, CA, USA) and quantified using Quantus™ Fluorometer (Promega, USA). Purified amplicons were pooled in equimolar amounts, and paired-end sequenced on an Illumina MiSeq PE300 platform (Illumina, San Diego, USA) according to the standard protocols by Majorbio Bio-Pharm Technology Co. Ltd. (Shanghai, China).

Serum iFABP and LPS Measurement

The blood samples of all mice were collected via Eyeball, and the serum was separated by centrifugation at 1, 000×g for 20 minutes. Then, ELISA technology was used to measure the iFABP and LPS, according to the manufacturer’s instructions. The commercial ELISA kits were obtained from Shanghai Enzyme-linked Biotechnology Co., Ltd (Shanghai, China) including iFABP (catalog number TW9968) and LPS (catalog number TW12543).

Fecal microbial transplantation

To establish the gut microbiota depletion mouse model, SPF mice were treated with a cocktail of antibiotics (1 mg/mL ampicillin, 1 mg/mL metronidazole, 1 mg/mL neomycin, and 0.5 mg/mL vancomycin) by oral gavage daily for 14 days. All mice were randomly separated into two groups, the CON-recipient group (n = 10) and the CL-recipient group (n = 10). Each group of mice was randomly housed in two cages with five mice in each cage. The fecal microbial transplantation procedure is shown in Figure 4a. Fecal samples from mice of the CON group and CL group were collected in sterile containers. Then 1 g fecal sample was suspended in 5 ml sterile PBS, followed by the vortexes, sedimentation, and filtrate with a 100-μm cell strainer. The suspension of the same group was mixed as microbiota donors and immediately administered to the mice by oral gavage. 200 μl of the supernatant containing fecal microbiota from either CON or CL donors was transferred to microbiota-depleted mice by oral gavage every day, for 14 days. All recipient mice were infected intranasally with 107 CFU MS after stopping FMT 2 days, and 3 days later of the infection, all mice were sacrificed, and the colonic contents and lung tissue samples were collected for further analysis.

Lung Histological Assessment

After the MS-infected mice were sacrificed, the lung tissues were perfused with sterile PBS and fixed in 4% paraformaldehyde for three days, followed by paraffin embedding. For histopathological analysis, 5μm sections were cut and stained using a standard H&E protocol. Motic EasyScan whole-slide scanner was used for scanning histological sections and images were analyzed using Matic DSAaaistant Lite.

Lung tissue RNA extraction and RNA sequencing

The lung tissue from each mouse was separated aseptically and quickly stored in liquid nitrogen for subsequent RNA extraction. Total RNA was extracted from lung tissue using TRIzol® reagent (Dingguo Changsheng Biotechnology Co., Ltd, Beijing, China) according to the manufacturer’s protocol. RNA quantity and quality were determined using the NanoDrop 2000 Spectrophotometer (Thermo Scientific, USA). The Agilent 2100 Bioanalyzer (Agilent Technologies, Santa Clara, CA, USA) was used to assess RNA integrity. RNA-seq libraries were prepared using VAHTS Universal V6 RNA-seq Library Prep Kit following the manufacturer’s recommendations. Then the paired-end RNA-seq libraries were sequenced with the Illumina HiSeq X Ten platform (2 × 150 bp read length) by OE Biotech Co., Ltd (Shanghai, China). After the quality control, clean reads performed bioinformatics analysis.

Quantification of gene expression using RT-qPCR

Total RNA was extracted from mouse lungs or A549 cells using the Trigol (Dingguo Changsheng Biotechnology Co., Ltd, Beijing, China) according to the manufacturer’s protocol, and was reversed transcribed using the kit from All-in-One Script RTpremix (Kryptoner Mei Co., Ltd, Zhengzhou, China). Then RT-qPCR was performed with TB Green Premix Ex Taq [(TaKaRa Biotechnology, China) to evaluate the amount of mRNA expression according to the manufacturer’s recommendations. Subsequently, PCR products were detected on a sequence detection system. The primer sequences of RT-qPCR were listed in Table S7 in this study. The relative gene expression levels were calculated using the 2 -ΔΔCt method.

Construction of Nos2 over-expression Plasmid and transfection into A549 cells

To enhance Nos2 expression levels in A549 cells, the recombinant plasmid Nos2-pcDNA3.1 was constructed. Based on the nucleotide sequences of Nos2 (GenBank accession No: NM_010927), the forward and reverse primers Nos2F/Nos2R (Table S7) were used for cloning the open reading frame and inserted into pcDNA3.1 expression vector. The insert orientation was confirmed by separate XhoI and BamHI digests followed by agarose gel electrophoresis. The Nos2-pcDNA3.1 plasmid was mixed with lipo 8000 (Beyotime, Shanghai, China) and then transfected into A549 cells via electroporation. 24 hours post-transfection, transfection efficiency was determined using RT-qPCR to ensure Nos2 over-expression in A549 cells.

Mycobacterium smegmatis infects in A549 cells

A549, a human alveolar basal epithelial cell, was used to assess the infection capacity of MS. A549 Cells were inoculated into 24-well plates with a density of 2[105 cells per well, and were cultured in RPMI1640 medium (Gibco Laboratories, USA) supplemented with 10% fetal bovine serum (Sangon Biotech Co., Ltd., Shanghai, China) and 100 U/mL penicillin/streptomycin (Beyotime, Shanghai, China) at 37 °C. Nos2-pcDNA3.1 was transfected into A549 cells when the cell density reached 70-80%. After 24 h, the supernatant was replaced with fresh medium containing MS and cultured for 6h, 12 h, and 24h. Then the wells were washed 3 times with PBS to remove unattached bacteria. Subsequently, A549 cells were lysed with 1ml 0.1% TritonX-100 (Shanghai Beyotime Biotechnology Co., Ltd., Shanghai, China) per well to prepare appropriate dilutions, which were plated on MiddleBrook7H10 Agar plates cultured for standard colony counts.

Detection of NO, ROS, and Defb1

Generation of NO in the A549 cells was detected by Nitric Oxide (NO) assay kit (Nanjing Jiancheng Biotechnology Research Institute Co., Ltd, Nanjing, china). Briefly, Cells were seeded into 24-well plates at a density of 2[105 cells per well, and Nos2-pcDNA3.1 was transfected into A549 cells when the cell density reached 70-80%. After 24 h, the cells were harvested by trypsinization and centrifuged at 1,000 rpm for 10 minutes. Then NO concentration was detected with the spectrophotometric method according to the manufacturer’s protocols. ROS levels were measured using the DCFH-DA. Briefly, after 24 h post-transfection of Nos2-pcDNA3.1, DCFH-DA was added to the culture medium at a final concentration of 10 μmol/mL, and incubated for 1 h at 37°C. The cells were washed twice with PBS and then stained with 10 μmol/mL DCFH-DA at 25 °C for 30 min in the dark room.

Images were acquired by a confocal microscope (Nikon, Japan). For ROS quantification, A549 cells were collected, rinsed twice with PBS, and suspended in 10 μmol/mL DCFH-DA for 30 min. After incubation, fluorescence was detected at 485 nm (excitation) and 530 nm (emission) using a microplate reader. All these analyses were conducted in three replicates. The expression of Defb1 was detected using RT-qPCR, and the primer sequences of RT-qPCR are listed in Table S7.

Bioinformatics analysis

For microbiome analysis, the bioinformatics data were analyzed using the Majorbio Cloud platform (https://cloud.majorbio.com). The alpha diversity at the genus level was assessed according to the Chao 1 and inverse Simpson index. The beta diversity was calculated by principal coordinate analysis (PCoA). A permutational analysis of variance was performed to assess the variation in the taxonomic structure of microbiota communities between groups. The Wilcoxon rank sum test was used to assess the different structures of microbiota communities between groups. LEfSe analyses were performed to compare different biomarkers between groups. The trans-kingdom network figures were built using the package igraph (version 1.2.6). For transcriptomic analysis, the bioinformatics data were analyzed using the OE Cloud platform (https://www.oebiotech.com). Differential expression analysis was performed using the DESeq2 q value < 0.05 and foldchange > 2 was set as the threshold for significantly differential expression genes (DEGs). Venn diagram and volcano plot of DEGs were performed to explore gene expression patterns. The functional enrichment of the above DEGs was conducted using the Gene Ontology database (GO) (http://www.geneontology.org/) and the Kyoto Encyclopedia of Genes and Genomes (KEGG) (http://www.genome.jp/kegg/).

Statistical analysis

All statistics were performed using GraphPad Prism 8.0. If the data followed a normal distribution, unpaired Student’s t-tests were used to compare various parameters between the two groups. If the data did not follow a normal distribution, a non-parametric Mann-Whitney U test, and Wilcoxon rank sum test were used to compare the results. One-way analysis of variance was used for three or more groups of data. The graphs were made with GraphPad Prism 8.0 or R package (version 3.6.2). P-values of <0.05 were set as a threshold for statistical significance.

Data Availability

The raw data sets of 16s RNA and ITS Sequencing are available in the Sequence Read Archive (SRA) of the National Center for Biotechnology Information (NCBI), the BioProject number PRJNA1091926 (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA1091926). The raw data of the transcriptome are available in the SRA of the NCBI, the BioProject number PRJNA1099882 (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA1099882).

Ethics approval and consent to participate

The experimental procedures used in this study were approved by the Animal Care and Use Committee of Xinxiang Medical University, China (No. EC-023-098).

Supporting information